Species association networks unveil scale effects in community assemblage mechanisms.

### 2025-06-10 DEPRECATED/PAPER_CODE_files/PAPER_CODE.qmd
rewrote the code to adjust for the changes in the v3 draft

```{r, loading libraries}
library(dplyr)
library(PLNmodels)
library(igraph)
```

@@

  if (remove.isolated) {
    G <- delete.vertices(G, which(degree(G) == 0))
  }
  plot(G, layout = layout)
  invisible(G)
}

>>>>

```{r, loading libraries}
library(dplyr)
# library(PLNmodels)
library(igraph)
```

@@

  if (remove.isolated) {
    G <- delete.vertices(G, which(degree(G) == 0))
  }
  plot(G, layout = layout, font.size = 8)
  invisible(G)
}

### 2025-06-10 PAPER_CODE.qmd
rewrote the code to adjust for the changes in the v3 draft

## Useful libraries & functions

```{r}
library(dplyr)
# library(PLNmodels)
library(igraph)
```

```{r}
plot_network = function(net             ,
                         type            ,
                         output          ,
                         edge.color      = c("#F8766D", "#00BFC4"),
                         remove.isolated = FALSE,
                         node.labels     = NULL,
                         layout          = layout_in_circle,
                         plot = TRUE) {

  if (output == "igraph") {

    G <-  graph_from_adjacency_matrix(net, mode = "undirected", weighted = TRUE, diag = FALSE)

    if (!is.null(node.labels)) {
      igraph::V(G)$label <- node.labels
    } else {
      igraph::V(G)$label <- colnames(net)
    }
    ## Nice nodes
    V.deg <- degree(G)/sum(degree(G))
    igraph::V(G)$label.cex <- V.deg / max(V.deg) + .5
    igraph::V(G)$size <- V.deg * 100
    igraph::V(G)$label.color <- rgb(0, 0, .2, .8)
    igraph::V(G)$frame.color <- NA
    ## Nice edges
    igraph::E(G)$color <- ifelse(igraph::E(G)$weight > 0, edge.color[1], edge.color[2])
    if (type == "support")
      igraph::E(G)$width <- abs(igraph::E(G)$weight)
    else
      igraph::E(G)$width <- 15*abs(igraph::E(G)$weight)

    if (remove.isolated) {
      G <- delete.vertices(G, which(degree(G) == 0))
    }
    if (plot) plot(G, layout = layout)
  }
  if (output == "corrplot") {
    if (plot) {
      if (ncol(net) > 100)
        colnames(net) <- rownames(net) <- rep(" ", ncol(net))
      G <- net
      diag(net) <- 0
      corrplot(as.matrix(net), method = "color", is.corr = FALSE, tl.pos = "td", cl.pos = "n", tl.cex = 0.5, type = "upper")
    } else  {
      G <- net
    }
  }
  invisible(G)
}

@@

## Loading & formatting data

### Regional dataset

```{r}
load("REGIONAL_DATASET_COUNTS.Rdata")
load("REGIONAL_DATASET_COVARIATES.Rdata")
```

```{r}
regional_data <- prepare_data(counts     = REGIONAL_DATASET_COUNTS,
                              covariates = REGIONAL_DATASET_COVARIATES)
```

### Northern Range dataset

```{r}
load("NORTHERN_RANGE_DATASET_COUNTS.Rdata")
load("NORTHERN_RANGE_DATASET_COVARIATES.Rdata")
```

```{r}
northern_range_data <- prepare_data(counts     = NORTHERN_RANGE_DATASET_COUNTS,
                                    covariates = NORTHERN_RANGE_DATASET_COVARIATES)
```

## Running the ZIPLN-network model

We fix the seed for reproducibility.

```{r}
set.seed(1)
```

### Regional dataset

```{r}
regional_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = regional_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")

@@

We select a penalty using stability selection process StARS. As it stability selection requires a long time to run, we memorize the corresponding penalty and later use it directly to access the model selected with StARS.

```{r}
# regional_model_StARS <- regional_model$getBestModel("StARS")
regional_model_StARS <- regional_model$getModel(0.06881528)  # Avec le code git
# regional_model_StARS <- regional_model$getModel(0.4222431) # Avec la librairie
```

### Northern Range dataset

```{r}
northern_range_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = northern_range_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")

@@

We select a penalty using stability selection process StARS. As it stability selection requires a long time to run, we memorize the corresponding penalty and later use it directly to access the model selected with StARS.

```{r}

# northern_range_model_StARS <- northern_range_model$getBestModel("StARS")
northern_range_model_StARS <- northern_range_model$getModel(0.09219056)
```

## Resulting networks

We want to place species the datasets share in the same order for readibility.

```{r}
regional_species       <- colnames(regional_data$Abundance)
northern_range_species <- colnames(northern_range_data$Abundance)
intersecting_species   <- intersect(northern_range_species, regional_species)
```

### Regional network

Sorting species names.

```{r}
regional_sorted_species <- c(intersecting_species, regional_species[! regional_species %in% intersecting_species])
```

Extracting the association network from the model.

```{r}
regional_network        <- as.matrix(regional_model_StARS$latent_network("partial_cor"))
regional_sorted_indices <- match(regional_sorted_species, colnames(regional_network))
regional_network        <- regional_network[regional_sorted_indices, regional_sorted_indices] 
```

```{r}
plot_network(regional_network,
             type            = "partial_cor",
             output          = "igraph",
             edge.color      = c("#F8766D", "#00BFC4"),
             remove.isolated = FALSE,
             node.labels     = NULL,
             layout          = layout_in_circle,
             plot            = TRUE)
```

### Northern Range network

Sorting species names.

```{r}
northern_range_sorted_species <- c(intersecting_species, northern_range_species[! northern_range_species %in% intersecting_species])
```

```{r}
northern_range_network        <- as.matrix(northern_range_model_StARS$latent_network("partial_cor"))
northern_range_sorted_indices <- match(northern_range_sorted_species, colnames(northern_range_network))
northern_range_network        <- northern_range_network[northern_range_sorted_indices, northern_range_sorted_indices] 
```

```{r}
plot_network(northern_range_network,
             type            = "partial_cor",
             output          = "igraph",
             edge.color      = c("#F8766D", "#00BFC4"),
             remove.isolated = FALSE,
             node.labels     = NULL,
             layout          = layout_in_circle,
             plot            = TRUE)
```

>>>>

# Useful libraries & functions

PLNmodels version should be 1.2.2 or higher.

```{r, loading libraries}
# library(PLNmodels)
library(Metrics)
library(pheatmap)
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(viridisLite)
```

# Loading & formatting data

## Data loading

```{r data-loading}
load("data/NORTHERN_RANGE_DATASET_COUNTS.Rdata")
NR_COUNTS <- NORTHERN_RANGE_DATASET_COUNTS %>% as_tibble()
NR_COUNTS <- NR_COUNTS[, -21]

load("data/NORTHERN_RANGE_DATASET_COVARIATES.Rdata")
NR_COVARIATES <- 
  NORTHERN_RANGE_DATASET_COVARIATES %>% as_tibble() %>% 
  rename(latitude = LATITUDE, 
         longitude = LONGITUDE, 
         long_lat = LONGLAT,
         coarse_gravel = coarse.gravel, # We do not include geographical coordinates here
         fine_gravel = fine.gravel, 
         leaf_litter = leaf.litter,
         time_step = TimeStep, year = YEAR, month = MONTH, 
         stream = STREAM, 
         disturbance = DISTURBANCE, # human activity / binary categorical variable
         altitude = ALTITUDE, 
         season = SEASON) %>% 
  mutate(
    # season = case_match(time_step,
    #   c(1, 5,  9, 13, 17) ~ "dry_start",
    #   c(2, 6, 10, 14, 18) ~ "dry_end",
    #   c(3, 7, 11, 15, 19) ~ "wet_start",
    #   c(4, 8, 12, 16    ) ~ "wet_end"
    # ),
    season = as.character(season),
    year = as.character(year),
    disturbance = ifelse(disturbance == 1, "yes", "no")
  ) %>% 
  ## flow is not reliable
  ## time_step + month are redundant with season + year
  ## site is redudant with disturbance + stream
  ## long_lat is the 1st PCA axis of latitude + longitude
  dplyr::select(-latitude, -longitude, - long_lat, -site, -time_step,
                -month, -flow) %>% # we remove geographical coordinates
  relocate(altitude, width, depth, volume, garbage,
           conductivity, O2, pH, temperature, turbidity, # site features
           coarse_gravel, fine_gravel, leaf_litter, cobble, sand, silt, boulders, canopy, # soil
           season, year, # sampling time
           stream, disturbance # categorical: place + human activity
         )
# turbidity: ordinal variable indicating the level of turbidity
# season:-> dry-start/dry-end, wet-start/wet-end, (january-> may: dry)
quali_ind <- seq(ncol(NR_COVARIATES) - 3, ncol(NR_COVARIATES))

@@

## Data visualization

```{r}
log_abundance <- log(1 + NR_DATA$Abundance)
log_abundance_sorted <- log_abundance[order(NR_DATA$stream), ]

color_palette <- paletteer_c("ggthemes::Orange-Gold", n = 30)
palette_colors <- paletteer::paletteer_c("ggthemes::Orange-Gold", n = 29)  # 29 instead of 30
color_palette <- c("white", palette_colors)
max_val <- max(log_abundance_sorted, na.rm = TRUE)

pheatmap(log_abundance_sorted,
         color = color_palette,
         cluster_rows = FALSE, cluster_cols = FALSE,
         annotation_row = data.frame(Stream = NR_DATA$stream),
         show_rownames = FALSE)
```
## Covariates preparation (PCA)

```{r pca}
nb_pc <- ncol(NR_COVARIATES) - 4
NR_PCA <-
  NR_COVARIATES %>%
  PCA(quali.sup = quali_ind, ncp = nb_pc, scale.unit = TRUE, graph = FALSE)

NR_PCA_SCORES <- setNames(data.frame(NR_PCA$ind$coord), paste0("PC",1:nb_pc))
```

```{r data-formatting}
NR_DATA <- prepare_data(
  counts     = NR_COUNTS,
  covariates = cbind(NR_COVARIATES, NR_PCA_SCORES)
  )
```

# Running PLN-network

```{r pln_network_pc1_pc2}
model_PLN_pc1_pc2  <- PLNnetwork(Abundance ~ PC1 + PC2,  data = NR_DATA,
                         control = PLNnetwork_param(min_ratio = 0.05,
                                                    penalize_diagonal = FALSE))

# model_PLN_pc1_pc2_StARS <- model_PLN_pc1_pc1$getBestModel("StARS", stability = 0.8)
model_PLN_pc1_pc2_StARS <- model_PLN_pc1_pc2$getModel(0.2444352)
print(model_PLN_pc1_pc2_StARS$BIC)

model_PLN_pc1_pc2_StARS$plot_network()

B_log_scale <- sign(model_PLN_pc1_pc2_StARS$model_par$B) * log1p(abs(model_PLN_pc1_pc2_StARS$model_par$B))
rownames(B_log_scale) <- c("Intercept", "PC1", "PC2")

pheatmap(B_log_scale,color = paletteer::paletteer_c("viridis::viridis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)
```

```{r pln_network_pc1_pc1_disturbance}
model_PLN_pc1_pc2_disturbance  <- PLNnetwork(Abundance ~ PC1 + PC2 + disturbance,  data = NR_DATA,
                         control = PLNnetwork_param(min_ratio = 0.05,
                                                    penalize_diagonal = FALSE))
# model_PLN_pc1_pc2_disturbance_StARS <- model_PLN_pc1_pc2_disturbance$getBestModel("StARS", stability = 0.8)
model_PLN_pc1_pc2_disturbance_StARS <- model_PLN_pc1_pc2_disturbance$getModel(0.2361045)
print(model_PLN_pc1_pc2_disturbance_StARS$BIC)

model_PLN_pc1_pc2_disturbance_StARS$plot_network()

B_log_scale <- sign(model_PLN_pc1_pc2_disturbance_StARS$model_par$B) * log1p(abs(model_PLN_pc1_pc2_disturbance_StARS$model_par$B))
rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "disturbance")
pheatmap(B_log_scale,
         color = paletteer::paletteer_c("viridis::viridis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)
```

```{r pln_network}
model_PLN  <- PLNnetwork(Abundance ~ PC1 + PC2 + stream + disturbance,  data = NR_DATA,
                         control = PLNnetwork_param(min_ratio = 0.05,
                                                    penalize_diagonal = FALSE))
# model_PLN_StARS <- model_PLN$getBestModel("StARS", stability = 0.8)
model_PLN_StARS <- model_PLN$getModel(0.06570186)
print(model_PLN_StARS$BIC)

model_PLN_StARS$plot_network()

B_log_scale <- sign(model_PLN_StARS$model_par$B) * log1p(abs(model_PLN_StARS$model_par$B))
rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "stream_Caura", "stream_Lopinot",
                           "stream_LowerAripo", "stream_Maracas",   "stream_Quare",
                           "stream_Turure", "stream_UpperAripo", "disturbance")
pheatmap(B_log_scale,
         color = paletteer::paletteer_c("viridis::viridis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)
```

# Running ZIPLN-network

```{r zipln_network_pc1_pc2 }
model_ZIPLN_pc1_pc2   <- ZIPLNnetwork(Abundance ~ PC1 + PC2 | 1,  data = NR_DATA,
                         control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                      penalize_diagonal = FALSE))
# model_ZIPLN_pc1_pc2_StARS <- model_ZIPLN_pc1_pc2$getBestModel("StARS", stability = 0.8)
model_ZIPLN_pc1_pc2_StARS <- model_ZIPLN_pc1_pc2$getModel(0.05578001)
print(model_ZIPLN_pc1_pc2_StARS$BIC)

model_ZIPLN_pc1_pc2_StARS$plot_network()

B_log_scale <- sign(model_ZIPLN_pc1_pc2_StARS$model_par$B) * log1p(abs(model_ZIPLN_pc1_pc2_StARS$model_par$B))
rownames(B_log_scale) <- c("Intercept", "PC1", "PC2")
pheatmap(B_log_scale,
         color = paletteer::paletteer_c("viridis::viridis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)

pheatmap(model_ZIPLN_pc1_pc2_StARS$model_par$Pi,
          color = paletteer::paletteer_c("viridis::cividis", 30),
          cluster_rows = FALSE, cluster_cols = FALSE)
```

```{r zipln_network_pc1_pc2_disturbance }
model_ZIPLN_pc1_pc2_disturbance   <- ZIPLNnetwork(Abundance ~ PC1 + PC2 + disturbance | 1,  data = NR_DATA,
                                                  control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                      penalize_diagonal = FALSE))
# model_ZIPLN_pc1_pc2_disturbance_StARS <- model_ZIPLN_pc1_pc2_disturbance$getBestModel("StARS", stability = 0.8)
model_ZIPLN_pc1_pc2_disturbance_StARS <- model_ZIPLN_pc1_pc2_disturbance$getModel(0.04622685)
print(model_ZIPLN_pc1_pc2_disturbance_StARS$BIC)

model_ZIPLN_pc1_pc2_disturbance_StARS$plot_network()

B_log_scale <- sign(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$B) * log1p(abs(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$B))
rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "disturbance")
pheatmap(B_log_scale,
         color = paletteer::paletteer_c("viridis::viridis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)

pheatmap(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$Pi,
         color = paletteer::paletteer_c("viridis::cividis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)
```

```{r zipln_network}
model_ZIPLN  <- ZIPLNnetwork(Abundance ~ PC1 + PC2 + disturbance | stream,  data = NR_DATA,
                         control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                      penalize_diagonal = FALSE))
# model_ZIPLN_StARS <- model_ZIPLN$getBestModel("StARS", stability = 0.8)
model_ZIPLN_StARS <- model_ZIPLN$getModel(0.03317969)
print(model_ZIPLN_StARS$BIC)

model_ZIPLN_StARS$plot_network()

B_log_scale <- sign(model_ZIPLN_StARS$model_par$B) * log1p(abs(model_ZIPLN_StARS$model_par$B))
rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "disturbance")
pheatmap(B_log_scale,
         color = paletteer::paletteer_c("viridis::viridis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)

pheatmap(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$Pi,
         color = paletteer::paletteer_c("viridis::cividis", 30),
         cluster_rows = FALSE, cluster_cols = FALSE)
```

```{r}
pi_proba <- model_ZIPLN_StARS$model_par$Pi 
pi_proba_df <- as.data.frame(pi_proba)

STREAM <- NR_DATA$stream
# Assume M is a matrix and V is a vector of same length as nrow(M)

# Convert to data frame

pi_proba_df$STREAM <- STREAM

# Keep one row per unique value of V
pi_proba_unique <- pi_proba_df %>%
  group_by(STREAM) %>%
  slice(1) %>%
  ungroup()

# Extract matrix part
pi_proba_final <- as.matrix(select(pi_proba_unique, -STREAM))

# Set row names as the values of V
rownames(pi_proba_final) <- pi_proba_unique$V

# Plot the heatmap

pheatmap(pi_proba_final,
         color = rev(paletteer::paletteer_c("viridis::cividis", 30)),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = TRUE)  # show V as row names

@@

# Cross-validation measures for PLN-network and ZIPLN-network
```{r cross-val-folds}
K = 5
set.seed(1)
folds    = split(sample(1:nrow(covar_data)), rep(1:K, length = nrow(covar_data)))
```

```{r pln-cross-val-function}
crossval <- function(formula, data, K = 5,
                     folds    = split(sample(1:nrow(data)),
                                      rep(1:K, length = nrow(data))),
                     method   = c("PLN", "ZIPLN")
  ) {
    method <- match.arg(method)
    Y     <- lapply(folds, function(fold) data$Abundance[fold, ])
    Y_hat <- lapply(folds, function(test_set) {
      # rep(0, length(test_set))
      train_set <- setdiff(1:nrow(data), test_set)
      model <- do.call(method, list(formula = eval(formula), data = data, subset = train_set))
      predict(model, data[test_set, ], type = "response")
    })
    TP_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] == 0 & Y_hat[[k]] < 0.1)}))
    TN_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] != 0 & Y_hat[[k]] > 0.1)}))
    FP_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] != 0 & Y_hat[[k]] < 0.1)}))
    FN_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] == 0 & Y_hat[[k]] > 0.1)}))
    precision <- unlist(lapply(1:K, function(k){TP_zero[[k]] / (TP_zero[[k]] + FP_zero[[k]])}))
    recall    <- unlist(lapply(1:K, function(k){TP_zero[[k]] / (TP_zero[[k]] + FN_zero[[k]])}))
    F1_score  <- unlist(lapply(1:K, function(k){TP_zero[[k]] / (TP_zero[[k]] + .5 * (FP_zero[[k]] + FN_zero) )}))

    # err <- map2_dbl(Y_hat, Y, function(y_hat, y) mean((y_hat - y)^2))
    err <- map2_dbl(Y_hat, Y, function(y_hat, y) Metrics::rmse(y_hat, y))
    attr(err, "folds") <- folds
    return(list("err" = err,
                "TP_zero" = TP_zero, "TN_zero" = TN_zero,
                "FP_zero" = FP_zero, "FN_zero" = FN_zero,
                "precision" = precision, "recall" = recall,
                "F1_score" = F1_score))
  }
```

## PLN-network cross-validation
```{r}
cv_pln_out      <- crossval(Abundance ~ 1 + PC1 + PC2 + stream + disturbance,
                         data = NR_DATA)
print(median(cv_pln_out$err))

cv_pln_out_pc1_pc2      <- crossval(Abundance ~ 1 + PC1 + PC2,
                                    data = NR_DATA, folds = folds)
print(median(cv_pln_out_pc1_pc2$err))

cv_pln_out_pc1_pc2_disturbance      <- crossval(Abundance ~ 1 + PC1 + PC2 + disturbance,
                                    data = NR_DATA, folds = folds)
print(median(cv_pln_out_pc1_pc2_disturbance$err))
```

## ZIPLN-network cross-validation
```{r}
cv_pln_out_ZI   <- crossval(Abundance ~ 1 + PC1 + PC2 + disturbance | stream ,
                            data = NR_DATA, method = "ZIPLN")
print(median(cv_pln_out_ZI$err))

cv_pln_out_ZI_pc1_pc2   <- crossval(Abundance ~ 1 + PC1 + PC2 | 1 ,
                            data = NR_DATA, method = "ZIPLN", folds = folds)
print(median(cv_pln_out_ZI_pc1_pc2$err))

cv_pln_out_ZI_pc1_pc2_disturbance   <- crossval(Abundance ~ 1 + PC1 + PC2 + disturbance | 1 ,
                                    data = NR_DATA, method = "ZIPLN", folds = folds)
print(median(cv_pln_out_ZI_pc1_pc2_disturbance$err))
```

# Baseline methods for comparisons

```{r baselines-data-formatting}
nb_pc <- ncol(NR_COVARIATES) - 4
NR_PCA <-
  NR_COVARIATES %>%
  PCA(quali.sup = quali_ind, ncp = nb_pc, scale.unit = TRUE, graph = FALSE)

NR_PCA_SCORES <- setNames(data.frame(NR_PCA$ind$coord), paste0("PC", 1:nb_pc))

covariates = cbind(NR_COVARIATES, NR_PCA_SCORES)
NR_DATA <- prepare_data(
  counts     = NR_COUNTS,
  covariates = covariates
)
covar_data <- NR_DATA %>% select(c("PC1", "PC2", "disturbance", "stream"))
```
## BIC

```{r get-BIC-function}
get_BIC <- function(formula, covar_data, counts_data,
                    method = c("poisson", "log_normal")){
  method <- match.arg(method)
  species   <- colnames(counts_data)

  models <- lapply(species, function(S){
                   Y    <- counts_data[, S]
                   data <- cbind(covar_data, Y)
                   colnames(data)[[length(colnames(data))]] <- "Abundance"
                   if(method == "poisson"){
                     glm(eval(formula), data = data,
                         family = poisson(link = "log"))
                   }else{
                     log_data <- data
                     log_data$Abundance <- log(1 + data$Abundance)
                     glm(eval(formula), data = log_data,
                         family = gaussian(link = "identity"))
                   }
  })

  total_logLik <- sum(sapply(models, logLik))
  total_k <- sum(sapply(models, function(m) length(coef(m))))
  n_obs <- length(models[[1]]$y)  # assuming all same length
  total_BIC <- 2 * total_logLik - total_k * log(n_obs)
  return(total_BIC)
}
```

```{r BIC-poisson}

BIC_poisson <- get_BIC("Abundance ~ PC1 + PC2 + disturbance + stream", covar_data, NR_COUNTS,
                       method = "poisson")
print(BIC_poisson)

BIC_poisson_pcs_dist <- get_BIC("Abundance ~ PC1 + PC2 + disturbance", covar_data, NR_COUNTS,
                       method = "poisson")
print(BIC_poisson_pcs_dist)

BIC_poisson_pcs <- get_BIC("Abundance ~ PC1 + PC2", covar_data, NR_COUNTS,
                       method = "poisson")
print(BIC_poisson_pcs)
```

```{r BIC-log-normal}

BIC_normal <- get_BIC("Abundance ~ PC1 + PC2 + disturbance + stream", covar_data, NR_COUNTS,
                       method = "log_normal")
print(BIC_poisson_pcs)

BIC_normal_pcs_dist <- get_BIC("Abundance ~ PC1 + PC2 + disturbance", covar_data, NR_COUNTS,
                       method = "log_normal")
print(BIC_normal_pcs_dist)

BIC_normal_pcs <- get_BIC("Abundance ~ PC1 + PC2", covar_data, NR_COUNTS,
                           method = "log_normal")
print(BIC_normal_pcs)
```

## Cross-validation

```{r one-species-cross-val}
crosspredict_one_species <- function(formula, folds, data,
                                     method = c("poisson", "log_normal")){
  method <- match.arg(method)

  res <- lapply(folds, function(test_set) {
                         train_set <- setdiff(1:nrow(data), test_set)
                         if(method == "poisson"){
                           tmp   <- data[train_set,]
                           model <- glm(eval(formula),
                                        data = tmp,
                                        family = poisson(link = "log"))
                           list(Y_hat = predict(model, data[test_set, ], type = "response"),
                                Y_fit = model$fitted.values)
                         }else{
                           log_data <- data
                           log_data$Abundance <- log(1 + data$Abundance)
                           tmp   <- log_data[train_set,]
                           model <- glm(eval(formula),
                                        data = tmp,
                                        family = gaussian(link = "identity"))
                           list(Y_hat = exp(predict(model, data[test_set, ], type = "response")) - 1,
                                Y_fit = model$fitted.values)
                         }
                         }
                  )
  Y_hat = lapply(res, function(x) x$Y_hat)
  Y_fit = lapply(res, function(x) x$Y_fit)
  return(list("Y_hat" = Y_hat, "Y_fit" = Y_fit))
}
```

```{r overall-cross-val-baseline}

cross_val_baseline <- function(formula, folds, covar_data, counts_data,
                      method = c("poisson", "log_normal")){
  method <- match.arg(method)

  K <- length(folds)
  all_Y_train <- lapply(1:5, f<- function(x) c())
  all_Y_test     <- lapply(1:5, f<- function(x) c())
  all_Y_fit <- lapply(1:5, f<- function(x) c())
  all_Y_hat <- lapply(1:5, f<- function(x) c())
  species   <- colnames(counts_data)

  for(S in species){
    Y    <- counts_data[, S]
    data <- cbind(covar_data, Y)
    colnames(data)[[length(colnames(data))]] <- "Abundance"
    res <- crosspredict_one_species(formula, folds, data, method)
    Y_hat <- res$Y_hat ; Y_fit <- res$Y_fit
    for(k in 1:K){
      train_set <- setdiff(1:nrow(counts_data), folds[[k]])
      all_Y_train[[k]] <- rbind(all_Y_train[[k]], data$Abundance[train_set])
      all_Y_test[[k]]  <- rbind(all_Y_test[[k]], data$Abundance[folds[[k]]])
      all_Y_fit[[k]]   <- rbind(all_Y_fit[[k]], Y_fit[[k]])
      all_Y_hat[[k]]   <- rbind(all_Y_hat[[k]], Y_hat[[k]])
      # all_Y_hat[[k]] <- rbind(all_Y_hat[[k]], rep(0, length(folds[[k]])))
    }
  }

  predict_err <- c() ; fit_err <- c()
  for(k in 1:K){
    predict_err <- c(predict_err, Metrics::rmse(all_Y_test[[k]], all_Y_hat[[k]] ))
    fit_err     <- c(fit_err, Metrics::rmse(all_Y_train[[k]], all_Y_fit[[k]] ))
  }
  return(list("predict_err" = predict_err,
              "fit_err"     = fit_err))
}

@@

```{r poisson-cross-val}
cv_poisson_glm <- cross_val_baseline("Abundance ~ PC1 + PC2 + disturbance + stream",
                            folds, covar_data, NR_COUNTS, "poisson")
print(median(cv_poisson_glm$fit_err))
print(median(cv_poisson_glm$predict_err))

cv_poisson_glm_pcs <- cross_val_baseline("Abundance ~ PC1 + PC2",
                            folds, covar_data, NR_COUNTS, "poisson")
print(median(cv_poisson_glm_pcs$fit_err))
print(median(cv_poisson_glm_pcs$predict_err))

cv_poisson_glm_pcs_disturbance <- cross_val_baseline("Abundance ~ PC1 + PC2 + disturbance",
                                folds, covar_data, NR_COUNTS, "poisson")
print(median(cv_poisson_glm_pcs_disturbance$fit_err))
print(median(cv_poisson_glm_pcs_disturbance$predict_err))

```{r log-normal-cross-val}
cv_log_normal  <- cross_val_baseline("Abundance ~ PC1 + PC2 + disturbance + stream",
                            folds, covar_data, NR_COUNTS, "log_normal")
print(median(cv_log_normal$fit_err))
print(median(cv_log_normal$predict_err))

cv_log_normal_pcs  <- cross_val_baseline("Abundance ~ PC1 + PC2",
                            folds, covar_data, NR_COUNTS, "log_normal")
print(median(cv_log_normal_pcs$fit_err))
print(median(cv_log_normal_pcs$predict_err))

cv_log_normal_pcs_disturbance  <- cross_val_baseline("Abundance ~ PC1 + PC2+ disturbance",
                                folds, covar_data, NR_COUNTS, "log_normal")
print(median(cv_log_normal_pcs_disturbance$fit_err))
print(median(cv_log_normal_pcs_disturbance$predict_err))
```
